Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No Photosynthetic Advantage

### 2023-10-16 ms/ms.tex
added autoref features to ms

\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers

@@

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a uniform
distribution of stomata on both abaxial and adaxial leaf surfaces and 2)
coordinated stomatal spacing on each surface that offsets the position
of stomata (Fig. \ref{fig:amphi_grid}). Coordination between leaf
surfaces is defined in this study as the occurrence of stomata in areas
farthest from stomata on the opposite leaf surface. Additionally,
because CO\(_2\) is more limiting for photosynthesis under high light,
we hypothesize that in high light 3) there should be more stomata, and
4) stomata should be more uniformly distributed than in low light.
Finally, as stomatal densities are selected for optimal operational
aperture, we hypothesize that 5) stomatal length will be positively
correlated with the area of the leaf surface to which it is closest. We
refer to this as the `stomatal zone', the leaf area surrounding a focal
stomate closest to that stomate and therefore the zone it supplies with
CO\(_2\)). This way, each stomate can be optimally sized relative to the
mesophyll volume it supplies.

To test these hypotheses, we grew the model plant \emph{Arabidopsis
thaliana} in high, medium, and low light and measured stomatal density,
size, and patterning on both leaf surfaces, and spatial coordination
between them. We use Voronoi tessellation techniques to calculate
stomatal zones. We also used a 2-D porous medium approximation of
CO\(_2\) diffusion and photosynthesis to predict the photosynthetic
advantage of optimal versus suboptimal coordination in stomatal
coordination between surfaces. Specifically, we predicted that traits
which affect diffusion path length (leaf thickness, stomatal density,
leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate
(temperature, pressure), and CO\(_2\) demand (Rubisco concentration,
light) would modulate the advantage of optimal stomatal arrangement
following the relationships outlined in Table \ref{tab:hypotheses}.
Here, we integrate over reasonable parameter space to determine the
ecophysiological context most likely to favor stomatal spatial
coordination in amphistomatous leaves.

\begin{table}[ht]
\centering
\begin{tabular}{rll}
  \hline
 & trait & relationship \\ 
  \hline
1 & leaf thickness & + \\ 
  2 & stomatal density & - \\ 
  3 & leaf porosity & - \\ 
  4 & lat.-vert. diffusion ratio & - \\ 
  5 & temperature & - \\ 
  6 & pressure & - \\ 
  7 & Rubisco concentration & + \\ 
  8 & light & + \\ 
   \hline
\end{tabular}
\caption{A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.} 
\label{tab:hypotheses}
\end{table}

@@

Finally, we examined the relationship between stomatal zone area and
stomatal length using a Bayesian generalized non-linear multilevel model
fit with the \emph{R} package \textbf{brms} version 2.15.0
(\protect\hyperlink{ref-burkner_brms_2017}{Bürkner 2017},
\protect\hyperlink{ref-burkner_advanced_2018}{2018}) and \emph{Stan}
version 2.27.0
(\protect\hyperlink{ref-stan_development_team_stan_2023}{\textbf{stan\_development\_team\_stan\_2023?}}).
Stomatal zone area was calculated using Voronoi tessellation (e.g.~Fig.
\ref{fig:example}). The stomatal zone area, \(S_\text{area}\), is the
region of the leaf surface whose distance to stomate, \(S\), is less
than the distance to any other stomate, \(S\). Stomatal length was
measured in ImageJ
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}).

@@

To test whether the position of ab- and adaxial stomata are coordinated
we compared the observed distribution to a null distribution where the
positions on each surface are random. For each pair of surfaces
(observed or synthetic) we calculated the distance squared between each
to the nearest stomatal centroid with the \emph{R} package
\textbf{raster} version 3.6.23. Then we calculated the cell-wise Pearson
correlation coefficient. If stomatal positions on each surface are
coordinated to minimize the distance between mesophyll and the nearest
stomate, then we expect a negative correlation. A cell that is far from
a stomate on one surface should be near a stomate on the other surface
(Fig. \ref{fig:amphi}). We generated a null distribution of the
correlation coefficient by simulating \(10^3\) synthetic data sets for
each observed pair. For each synthetic data set, we simulated stomatal
position using a random uniform distribution, as described above,
matching the number of stomata on abaxial and adaxial leaf surfaces.
Stomatal positions on each surface are coordinated if the correlation
coefficient is greater than 95\% of the synthetic correlation values
(one-tailed test).

@@

We modeled the coordination advantage over a range of leaf thicknesses,
stomatal densities, photosynthetic capacities, and light environments to
understand when offsetting stomatal position on each surface might
deliver a significant photosynthetic advantage (Table
\ref{tab:model_var}).

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
Variable & Parameter range & Units \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(I_0\) & \(50-1000\) & \(\mu\)mol m\(^{-2}\) s\(^{-1}\) \\
\(\varphi_\text{pal}\) & \(0.1-0.3\) & m\(^3\) airspace m\(^{-3}\)
leaf \\
\(T_\text{leaf}\) & \(101-501\) & \(\mu\)m \\
\(U\) & \(17-169\) & \(\mu\)m \\
\end{longtable}

WORKING HERE - make table of parameter combinations!

@@

If our hypotheses that natural selection will act to reduce the
diffusion path length of CO\(_2\) while minimizing stomatal number are
correct, then we would expect to find leaf surfaces with uniformly
distributed stomata. However, to the contrary, we find that though 57 of
the 132 (43.1\%) leaf surfaces were significantly more uniformly
dispersed than uniform random synthetic stomatal grids (\(\alpha\) =
0.05), none of the leaf surfaces exhibited perfectly uniform stomatal
patterning (dispersion index = 1) (Fig. \ref{fig:dispersion}).
Additionally, we hypothesized that as CO\(_2\) is more limiting to
photosynthesis under high light, stomata would be more uniformly
dispersed in plants grown in high light than plants grown in low and
medium light. The data also fail to support this hypothesis as there is
no strong, discernible trend between light and stomatal patterning.
Interestingly, adaxial leaf surfaces were more uniformly dispersed than
associated abaxial leaf surfaces across all light treatments
(F\(_\text{1,126}\) = 28.8; p-value \textless{} 0.001). Rather than
regulate stomatal patterning in response to light regimes, plants
respond by increasing stomatal density (Fig. \ref{fig:density}).
Stomatal density drastically increased in plants grown under high light,
validating the long held hypothesis that light strongly influences
stomatal density (F\(_\text{2,126}\) = 680.7; p-value \textless{}
0.001).

Across all light treatments and leaf surfaces, stomatal length and
stomatal area were weakly positively correlated, indicating some support
for our hypothesis that stomata are selected for an optimal operational
aperture (Fig. \ref{fig:length-area}).

@@

As evidence against our hypothesis that natural selection should favor
spatial coordination in the placement of stomata between abaxial and
adaxial leaf surfaces, we found no correlation between paired abaxial
and adaxial leaf surfaces (Fig. \ref{fig:dual-surface}). Light treatment
had no effect on correlation between surfaces (F\(_\text{2,63}\) = 2.28;
p-value = 0.11). All but one abaxial, adaxial surface pairs were
independent (\(\alpha\) = 0.05).

@@

\hypertarget{draft-text}{%
\subsubsection{DRAFT TEXT}\label{draft-text}}

During a model run, CO\(_2\) flux between neighboring nodes was
determined according to the mesophyll CO\(_2\) concentration (\(C_m\))
in each node, distance between cells, temperature, pressure, and leaf
porosity. Assimilation rate (\(A\)) in each node was determined by
\(A_\text{max}\), \(c_i\), irradiance, temperature, chlorophyll volume
concentration, Rubisco octamer concentration per mol chlorophyll, and
maximal oxygenation and carboxylation rates of Rubisco. Stomata were
placed either 1) directly on top of one another to represent
anticoordination or 2) offset from one another to represent coordination
between leaf surfaces. \(c_i\) and CO\(_2\) assimilation rate (A) were
set to 415ppm and 0 respectively and the model ran until \(c_i\) and A
reached equilibrium in each cell. Total A and average \(c_i\) were then
calculated for each parameter combination.

A was calculated in terms of \(A_\text{max}\) and \(c_i\) following Eq.
\ref{eq:eq4}

\begin{equation}\label{eq:eq4}
A = A_\text{max} * c_i / (c_i + K_c)
\end{equation}

where \(K_c\) is the Michaelis constant for CO\(_2\): 0.0184 mol
\(m^\text{-3}\).

\(A_\text{max}\) was calculated in terms of irradiance (i), light
saturated maximum A (\(A_\text{mm}\)), and quantum yield (\(\varphi\))
following Eq. \ref{eq:eq5}

\begin{equation}\label{eq:eq5}
A_\text{max} = A_\text{mm} * \varphi * i / (A_\text{mm} + \varphi * i)
\end{equation}

@@

Stomata are expensive. A theoretical, optimized plant would minimize
stomatal density while also allowing competitive gas exchange rates for
its environment so as to maximize C assimilation per unit investment in
stomata. Natural selection operates within developmental and physical
constraints to drive each plant species toward its theoretical optimum.
This study provides evidence that stomata in \emph{Arbidopsis thaliana},
the model angiosperm, are non-randomly distributed, favoring dispersion
over clustering (Fig. \ref{fig:dispersion}). However, stomata are not
ideally dispersed in an equilateral triangular grid as would be optimal
to minimize CO\(_2\) diffusion path length and standardize the area
supplied by each stomate (Fig. \ref{fig:example}). Additionally, when
grown in high light environments, \emph{A. thaliana} exhibited increased
stomatal density rather than increased stomatal dispersion (Fig.
\ref{fig:density}), which suggests that natural selection has acted more
strongly on developmental pathways that modulate stomatal density than
those that control stomatal dispersion. In other words, plants optimize
gas exchange by adding more stomata rather than dispersing them more
evenly across the leaf surface. This study also demonstrates that
stomata that supply larger leaf areas with CO\(_2\) tend to be larger
(Fig. \ref{fig:length-area}). These results could suggest that 1) the
added energetic and hydraulic cost of non-ideally dispersed stomata is
negligible and therefore not acted on by natural selection; 2) no
developmental pathway exists to ensure the ideal placement of stomata on
the leaf; or 3) the regulation of stomatal size limits the cost incurred
by non-ideal stomatal dispersion.

In high light environments, amphistomy is favorable as high light
photosynthesis is limited by CO\(_2\) and amphistomy halves diffusion
path length and boundary layer resistance, thereby reducing CO\(_2\)
limitation - increasing theoretical \(A_\text{max}\). An optimal
amphistomatous leaf has offset stomata such that stomata are more likely
to appear on one leaf surface if there is not a stomata directly
opposite it on the other surface as shown in figure \ref{fig:amphi}.
However, our results show that leaf surfaces are not coordinated but are
independent, regardless of light (Fig. \ref{fig:dual-surface}).
Additionally, gas exchange models show little photosynthetic efficiency
gain from abaxial-adaxial stomatal coordination compared to
anticoordination (INSERT FIG FROM MODELING). We posit that this marginal
gain is not sufficient to be acted upon strongly by natural selection.
Thus, amphistomatous plants do not exhibit abaxial-adaxial stomatal
coordination for there is little selective advantage of it.

@@

Beyond disperson on a single surface, gas exchange can be optimized via
stomatal coordination of abaxial and adaxial surfaces in amphistomatous
leaves. Given that leaf thicknesses are generally multiple times greater
than interstomatal distance (GIVE DISTANCES HERE). As a result,
abaxial-adaxial stomatal coordination reduces CO\(_2\) diffusion path
length far less than single surface dispersion, so we hypothesize this
strategy to afford less photosynthetic advantage to the leaf. Our
modelling results demonstrate that, even in ideal conditions, i.e.~thick
leaf, low stomatal densities, high light, low leaf porosity, high
rubisco concentration, etc., the photosynthetic advantage of
coordination is minimal. We are not surprised by these results, but
still highlight them here as we are the first to report this finding.

>>>>

\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}

@@

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a uniform
distribution of stomata on both abaxial and adaxial leaf surfaces and 2)
coordinated stomatal spacing on each surface that offsets the position
of stomata (\autoref{fig:ideal-amphi-grid}). Coordination between leaf
surfaces is defined in this study as the occurrence of stomata in areas
farthest from stomata on the opposite leaf surface. Additionally,
because CO\(_2\) is more limiting for photosynthesis under high light,
we hypothesize that in high light 3) there should be more stomata, and
4) stomata should be more uniformly distributed than in low light.
Finally, as stomatal densities are selected for optimal operational
aperture, we hypothesize that 5) stomatal length will be positively
correlated with the area of the leaf surface to which it is closest. We
refer to this as the `stomatal zone', the leaf area surrounding a focal
stomate closest to that stomate and therefore the zone it supplies with
CO\(_2\)). This way, each stomate can be optimally sized relative to the
mesophyll volume it supplies.

To test these hypotheses, we grew the model plant \emph{Arabidopsis
thaliana} in high, medium, and low light and measured stomatal density,
size, and patterning on both leaf surfaces, and spatial coordination
between them. We use Voronoi tessellation techniques to calculate
stomatal zones. We also used a 2-D porous medium approximation of
CO\(_2\) diffusion and photosynthesis to predict the photosynthetic
advantage of optimal versus suboptimal coordination in stomatal
coordination between surfaces. Specifically, we predicted that traits
which affect diffusion path length (leaf thickness, stomatal density,
leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate
(temperature, pressure), and CO\(_2\) demand (Rubisco concentration,
light) would modulate the advantage of optimal stomatal arrangement
following the relationships outlined in \autoref{tab:hypotheses}. Here,
we integrate over reasonable parameter space to determine the
ecophysiological context most likely to favor stomatal spatial
coordination in amphistomatous leaves.

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \hline
trait & relationship \\ 
  \hline
leaf thickness & + \\ 
  stomatal density & - \\ 
  leaf porosity & - \\ 
  lat.-vert. diffusion ratio & - \\ 
  temperature & - \\ 
  pressure & - \\ 
  Rubisco concentration & + \\ 
  light & + \\ 
   \hline
\end{tabular}
\caption{A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.} 
\label{tab:hypotheses}
\end{table}

@@

Finally, we examined the relationship between stomatal zone area and
stomatal length using a Bayesian generalized non-linear multilevel model
fit with the \emph{R} package \textbf{brms} version 2.15.0
(\protect\hyperlink{ref-burkner_brms_2017}{Bürkner 2017},
\protect\hyperlink{ref-burkner_advanced_2018}{2018}) and \emph{Stan}
version 2.27.0
(\protect\hyperlink{ref-stan_development_team_stan_2023}{\textbf{stan\_development\_team\_stan\_2023?}}).
Stomatal zone area was calculated using Voronoi tessellation
(e.g.~\autoref{fig:tessellation-example}). The stomatal zone area,
\(S_\text{area}\), is the region of the leaf surface whose distance to
stomate, \(S\), is less than the distance to any other stomate, \(S\).
Stomatal length was measured in ImageJ
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}).

@@

To test whether the position of ab- and adaxial stomata are coordinated
we compared the observed distribution to a null distribution where the
positions on each surface are random. For each pair of surfaces
(observed or synthetic) we calculated the distance squared between each
to the nearest stomatal centroid with the \emph{R} package
\textbf{raster} version 3.6.23. Then we calculated the cell-wise Pearson
correlation coefficient. If stomatal positions on each surface are
coordinated to minimize the distance between mesophyll and the nearest
stomate, then we expect a negative correlation. A cell that is far from
a stomate on one surface should be near a stomate on the other surface
(\autoref{fig:ideal-amphi-grid}). We generated a null distribution of
the correlation coefficient by simulating \(10^3\) synthetic data sets
for each observed pair. For each synthetic data set, we simulated
stomatal position using a random uniform distribution, as described
above, matching the number of stomata on abaxial and adaxial leaf
surfaces. Stomatal positions on each surface are coordinated if the
correlation coefficient is greater than 95\% of the synthetic
correlation values (one-tailed test).

@@

We modeled the coordination advantage over a range of leaf thicknesses,
stomatal densities, photosynthetic capacities, and light environments to
understand when offsetting stomatal position on each surface might
deliver a significant photosynthetic advantage
(\autoref{tab:model_var}).

\begin{table}

\caption{\label{tab:model_var}The parameter range of model variables tested for their effect on coordination advantage (\autoref{eq:coordination_advantage}) using a 2-D porous medium approximation. We used regularly spaced values within each range and simulated across all combinations. Here we converted model units to more conventional units (e.g. m to $\mu$m). $I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance}
\centering
\begin{tabular}[t]{lll}
\toprule
Variable & Parameter range & Units\\
\midrule
$I_0$ & $50-1000$ & $\mu$mol m$^{-2}$ s$^{-1}$\\
$\varphi_\text{pal}$ & $0.1-0.3$ & m$^3$ airspace m$^{-3}$ leaf\\
$T_\text{leaf}$ & $101-501$ & $\mu$m\\
$U$ & $17-169$ & $\mu$m\\
\bottomrule
\end{tabular}
\end{table}

WORKING HERE - ???

@@

If our hypotheses that natural selection will act to reduce the
diffusion path length of CO\(_2\) while minimizing stomatal number are
correct, then we would expect to find leaf surfaces with uniformly
distributed stomata. However, to the contrary, we find that though 57 of
the 132 (43.1\%) leaf surfaces were significantly more uniformly
dispersed than uniform random synthetic stomatal grids (\(\alpha\) =
0.05), none of the leaf surfaces exhibited perfectly uniform stomatal
patterning (dispersion index = 1) (\autoref{fig:single-surface}).
Additionally, we hypothesized that as CO\(_2\) is more limiting to
photosynthesis under high light, stomata would be more uniformly
dispersed in plants grown in high light than plants grown in low and
medium light. The data also fail to support this hypothesis as there is
no strong, discernible trend between light and stomatal patterning.
Interestingly, adaxial leaf surfaces were more uniformly dispersed than
associated abaxial leaf surfaces across all light treatments
(F\(_\text{1,126}\) = 28.8; p-value \textless{} 0.001). Rather than
regulate stomatal patterning in response to light regimes, plants
respond by increasing stomatal density (\autoref{fig:density}). Stomatal
density drastically increased in plants grown under high light,
validating the long held hypothesis that light strongly influences
stomatal density (F\(_\text{2,126}\) = 680.7; p-value \textless{}
0.001).

Across all light treatments and leaf surfaces, stomatal length and
stomatal area were weakly positively correlated, indicating some support
for our hypothesis that stomata are selected for an optimal operational
aperture (\autoref{fig:length-area}).

@@

As evidence against our hypothesis that natural selection should favor
spatial coordination in the placement of stomata between abaxial and
adaxial leaf surfaces, we found no correlation between paired abaxial
and adaxial leaf surfaces (\autoref{fig:dual-surface}). Light treatment
had no effect on correlation between surfaces (F\(_\text{2,63}\) = 2.28;
p-value = 0.11). All but one abaxial, adaxial surface pairs were
independent (\(\alpha\) = 0.05).

@@

@@

Stomata are expensive. A theoretical, optimized plant would minimize
stomatal density while also allowing competitive gas exchange rates for
its environment so as to maximize C assimilation per unit investment in
stomata. Natural selection operates within developmental and physical
constraints to drive each plant species toward its theoretical optimum.
This study provides evidence that stomata in \emph{Arbidopsis thaliana},
the model angiosperm, are non-randomly distributed, favoring dispersion
over clustering (\autoref{fig:single-surface}). However, stomata are not
ideally dispersed in an equilateral triangular grid as would be optimal
to minimize CO\(_2\) diffusion path length and standardize the area
supplied by each stomate (\autoref{fig:tessellation-example}).
Additionally, when grown in high light environments, \emph{A. thaliana}
exhibited increased stomatal density rather than increased stomatal
dispersion (\autoref{fig:density}), which suggests that natural
selection has acted more strongly on developmental pathways that
modulate stomatal density than those that control stomatal dispersion.
In other words, plants optimize gas exchange by adding more stomata
rather than dispersing them more evenly across the leaf surface. This
study also demonstrates that stomata that supply larger leaf areas with
CO\(_2\) tend to be larger (\autoref{fig:length-area}). These results
could suggest that 1) the added energetic and hydraulic cost of
non-ideally dispersed stomata is negligible and therefore not acted on
by natural selection; 2) no developmental pathway exists to ensure the
ideal placement of stomata on the leaf; or 3) the regulation of stomatal
size limits the cost incurred by non-ideal stomatal dispersion.

In high light environments, amphistomy is favorable as high light
photosynthesis is limited by CO\(_2\) and amphistomy halves diffusion
path length and boundary layer resistance, thereby reducing CO\(_2\)
limitation - increasing theoretical \(A_\text{max}\). An optimal
amphistomatous leaf has offset stomata such that stomata are more likely
to appear on one leaf surface if there is not a stomata directly
opposite it on the other surface as shown in
\autoref{fig:ideal-amphi-grid}. However, our results show that leaf
surfaces are not coordinated but are independent, regardless of light
(\autoref{fig:dual-surface}). Additionally, gas exchange models show
little photosynthetic efficiency gain from abaxial-adaxial stomatal
coordination compared to anticoordination (INSERT FIG FROM MODELING). We
posit that this marginal gain is not sufficient to be acted upon
strongly by natural selection. Thus, amphistomatous plants do not
exhibit abaxial-adaxial stomatal coordination for there is little
selective advantage of it.

@@

Beyond disperson on a single surface, gas exchange can be optimized via
stomatal coordination of abaxial and adaxial surfaces in amphistomatous
leaves. Given that leaf thicknesses are generally multiple times greater
than interstomatal distance (GIVE DISTANCES HERE). As a result,
abaxial-adaxial stomatal coordination reduces CO\(_2\) diffusion path
length far less than single surface dispersion, so we hypothesize this
strategy to afford less photosynthetic advantage to the leaf. Our
modelling results demonstrate that, even in ideal conditions, i.e.~thick
leaf, low stomatal densities, high light, low leaf porosity, high
Rubisco concentration, etc., the photosynthetic advantage of
coordination is minimal. We are not surprised by these results, but
still highlight them here as we are the first to report this finding.
